Photo-D with LSST: stellar photometric distances out to the edge of the Galaxy

### 2024-05-28 paper/methodology.tex
comments from coauthors

In most general terms, our aim is to estimate for each star an array (a vector) of model parameters $\vec{\theta}$,
for some model $M$ (e.g., main-sequence stars), using data vector $\vec{D}$ and priors for $M$ and $\vec{\theta}$.
Data, or observations, include multi-band photometry that is used
to construct colors, $\vec{c}$. In the case of SDSS, colors include $u-g$, $g-r$, $r-i$, $i-z$, and in the case of LSST also
$z-y$. In addition to colors, $\vec{D}$ also includes an apparent magnitude and hereafter we choose the $r$-band
magnitude. Therefore, $\vec{D}$ = ($r$, $\vec{c}$). Observations also provide stellar sky coordinates and we
address their role further below when discussing priors.

@@

The observed stellar colors also depend on the interstellar dust extinction along the line of sight. Hereafter
we will assume (and justify further below) that dust extinction is fully specified by a single model parameter, $A_r$.
$A_r$ is extinction in the $r$ band and extinction in other bands is proportional to $A_r$, with known constants
of proportionality (this assumption can be relaxed, see below). Once $M_r$ and $A_r$ are constrained, the distance
modulus $\Delta$ can be computed from
\begin{equation}
  \label{eq:distmod}
                   r = M_r + A_r + \Delta,
\end{equation}
or in a probabilistic form
\begin{equation}
  \label{eq:distmodpdf}
                       p(\Delta) = p(r) - p(M_r + A_r). 
\end{equation}
Note that in practice the uncertainty of the observed magnitude, $r^{obs}$, is always significantly smaller than
the width of $p(M_r + A_r)$ distribution (at the bright end by an order of magnitude, $\sim$0.01 mag vs. $\sim$0.1 mag).

@@

\begin{figure*}[t!]
\includegraphics[width=1.0\textwidth,angle=0]{figures/plot3diagsData.png} 
\caption{An illustration of multiple stellar populations. The dots (of
any color) in the left panel show color-magnitude diagram for 841,000 stars from the SDSS Stripe 82 Standard Star Catalog
(variable sources are excluded, \citealt{2021MNRAS.505.5941T}) that have Gaia matches within 0.15 arcsec (after correcting for proper motion using Gaia measurements). A subset of 415,000 stars with $r < 22$ and $u<22$ are overplotted as blue dots, and 409,000 of those that also have $0.2 < g-i < 3.5$ (dominated by main-sequence stars and red giants) are overplotted as cyan dots. Finally, 63,000 stars that have signal-to-noise ratio for Gaia’s parallax measurements of at least 20 are shown as green dots (these stars can be used for the calibration of luminosity-color relations). The same symbol color scheme is used in other two panels. The three yellow lines in the middle panel show stellar locus parametrization used by Green et al. (2014) for three values of metallicity (left to right): $[Fe/H] = -2, -1, 0$. In the right panel, the impact of metallicity on color-color tracks is negligible and all three are indistinguishable from each other. In the bottom of the middle panel, at $-0.5 < g-r < 0$, the three dark blue sequences correspond to (from left to right) He white dwarfs, H white dwarfs, and blue horizontal branch stars. The clouds of pale blue dots visible above the main stellar locus in the middle and right panels correspond to unresolved binary stars \citep{2004ApJ...615L.141S}.} \label{fig:3dataDiags}
\end{figure*}

Our approach adopted here is essentially the same as used in recent papers\footnote{This approach greatly simplifies when studying
faint and distant blue halo stars, as in e.g., \cite{2008ApJ...673..864J}. Such stars are beyond the dust layer which is confined close
to the disk and thus $A_r$ can be obtained from IR maps, they have halo metallicities ($[Fe/H] \sim -1.5$), and can be assumed dominated
by main-sequence stars. As a result, a simple functional relationship, $M_r = f(g-i)$, or its generalized version that
accounts for the shift of $M_r$ as a function of metallicity \citep{2008ApJ...684..287I}, can be used to estimate distance
in a straightforward manner.}
by, e.g., \cite{2011MNRAS.411..435B},  \cite{2014ApJ...783..114G}, \cite{lallement_3d_2014}, 
\cite{gordon_panchromatic_2016}, \cite{queiroz_starhorse_2018}, \cite{green_3d_2019} and \cite{bailer-jones_estimating_2021}.
The main differences compared to these works include: 
\begin{enumerate}
\item The use of multiple stellar populations, in addition to main-sequence stars and red giants (white dwarfs, unresolved binary stars,
                blue horizontal branch stars; potentially Miras, quasars and RR Lyrae stars too, which can also be recognized and rejected using variability).
\item Improved color tracks for main-sequence stars and (especially) red giants, including the use of very young ($<$1 Gyr) populations 
               and an extended $[Fe/H]$ range. 
\item Priors based on sophisticated TRILEGAL simulations by \cite{2022ApJS..262...22D} that include multiple stellar populations and
               also account for the Galaxy's bulge component. 
\end{enumerate}

@@

\begin{figure*}[b!]
\includegraphics[width=1.0\textwidth,angle=0]{figures/plot3HRdiags.png}  
\caption{The left panel shows SDSS-based empirical absolute magnitude vs. color parametrization for main-sequence stars and red giants.
The data are color-coded by metallicity, ranging from $[Fe/H] = -2.5$ to 0 (blue to red). The three lines correspond to three values of
metallicity: $[Fe/H] = -2, -1, 0$ (dot-dashed, solid and dashed, respectively). The middle panel shows a sample of 63,000 stars that have
signal-to-noise ratio for Gaia’s parallax measurements of at least 20 (white dwarfs can be seen in the lower left corner.
Their absolute magnitudes are derived from parallax measurements. The dot-dashed,
solid and dashed black lines are the same as in the left panel. For comparison, the essentially identical yellow dotted lines were computed
using eqs. A2 and A7 from Ivezi\'{c} et al. (2008).  Note the discrepancy between these parametrizations and data for sub-giant stars ($M_r \sim 3-4$
and $g-i \sim 0.8-1.1$). The right panel shows a sample of 415,000 stars with $r < 22$ and $u<22$ as red dots (shown by blue dots in
Figure~\ref{fig:3dataDiags}), and 409,000 of those that also have  $0.2 < g-i < 3.5$ as cyan dots. Their absolute magnitudes were computed
using the so-called “photogeometric” distances from Bailer-Jones et al. (2021). The dot-dashed, solid and dashed black lines are the
same as in the left and middle panels. About 10,000 outliers (about 2.5\% of the full
sample) seen at $g-i = 0.4$ and $Mr > 7$ are predominantly found at the faint end ($r>20$).}
\label{fig:3HRdiags}
\end{figure*}

@@

We ``augment'' the empirical SDSS-based isochrones in two steps:
\begin{enumerate}
\item We extend its metallicity range from $[Fe/H]=0$ to $[Fe/H]=+0.5$ by linear extrapolation of the color vs. metallicity
  dependence around $[Fe/H]=0$. We adjust the gradient by a multiplicative factor (0.7) to ensure that the bright end of the
  stellar locus at $[Fe/H]=+0.5$ agrees with SDSS-Gaia data (shown in the middle panel in Figure~\ref{fig:3HRdiags}).
\item 
  For $M_r < M_{TO}$, where turn-off absolute magnitude $M_{TO}$ depends on age and ranges from $M_{TO}$=4 for age = 1 Gyr
  to $M_{TO}$ = 5 for age = 10 Gyr, we ``attach'' model-based isochrones (we use DSED isochrones hereafter) to empirical
  SDSS-based isochrones.
\end{enumerate}

@@

We sample the M dwarf luminosity in the range $8.5 \le M_r \le 14.5$, with a step of 0.1 mag, and for each value
generate a track by adding luminosity of a white dwarf, where the track is sampled on the same grid of $M_r$ but
using white dwarf models (separately for H and He models). There are four model families (for two families of white dwarf
models and two values of $[Fe/H]$ for M dwarfs), each with 3,721 $M_r$ vs. color entries. The color tracks for H white dwarfs
and $[Fe/H]=-1.5$ M dwarfs are shown in Figure~\ref{fig:locusBinaries}. Analogous tracks for binaries composed of an M dwarf with
$[Fe/H]=0$ and either an H or He white dwarf look very similar though not identical (differences are at most a few tenths
of a magnitude). In practice, it will be nearly impossible to constrain M dwarf metallicity, and very hard to distinguish
H and He white dwarfs. It may turn out that it will be sufficient to use a single model family to fit LSST data (e.g., a
combination of an H white dwarf and a $[Fe/H]=-1.5$ M dwarf).

@@

Recent studies, such as \cite{2012ApJ...757..166B} based on SDSS data (including the Galactic plane), 
find that the scatter of $R_V$ around its mean value $R_V=3.1$ is very small. For this reason, we adopt
empirical results from their Table 1: $C_m$ = (1.810, 1.400, 0.759, 0.561) in $(u, g, i, z)$, respectively ($C_r=1$
by definition). This choice does not preclude the use of our framework to fit for $R_V$ as the fourth free
model parameter in Galactic plane regions with very large $A_r$ ($R_V$ is poorly constrained when $A_r$ is small);
we  can simply add multiple models with extincted colors generated using different values of $C_m$.

@@

In the next Section, we discuss a fast numerical pipeline implementation that can perform this computation for LSST-sized
catalogs and provide a more quantitative analysis of the method's performance.  We conclude this Section with a brief
discussion of Bayesian model selection method for assigning posterior probabilities to each stellar population used to interpret
observations.

>>>>

In the most general terms, our aim is to estimate for each star an array (a vector) of model parameters $\vec{\theta}$,
for some model $M$ (e.g., main-sequence stars), using data vector $\vec{D}$ and priors for $M$ and $\vec{\theta}$.
Data, or observations, include multi-band photometry that is used
to construct colors, $\vec{c}$. In the case of SDSS, colors include $u-g$, $g-r$, $r-i$, $i-z$, and in the case of LSST also
$z-y$. In addition to colors, $\vec{D}$ also includes an apparent magnitude and hereafter we choose the $r$-band
magnitude. Therefore, $\vec{D}$ = ($r$, $\vec{c}$). Observations also provide stellar sky coordinates and we
address their role further below when discussing priors.

@@

The observed stellar colors also depend on the interstellar dust extinction along the line of sight. Hereafter
we will assume (and justify further below) that dust extinction is fully specified by a single model parameter, $A_r$.
$A_r$ is extinction in the $r$ band and extinction in other bands is proportional to $A_r$, with known constants
of proportionality (this assumption can be relaxed, see below). Once $M_r$ and $A_r$ are constrained, that is, the
posterior probability distribution $q(Q_r)$, where $Q_r=M_r+A_r$ is known, the distance modulus $\Delta$ can be
computed using relationship
\begin{equation}
  \label{eq:distmod}
                 r = M_r + A_r + \Delta.
\end{equation}
In practice the uncertainty of the observed magnitude, $r^{obs}$, is always much smaller than the width of $q(Q_r)$
distribution (at the bright end by an order of magnitude, $\sim$0.01 mag vs. $\sim$0.1 mag). Therefore,
\begin{equation}
  \label{eq:distmodpdf}
                         p(\Delta) = q(r^{obs} - \Delta),
\end{equation}
with the mean values related as $\mean{\Delta}$ = $r^{obs} - \mean{Q_r}$ and the uncertainty of $\Delta$
approximately equal to the standard deviation of $q(Q_r)$.

@@

\begin{figure*}[t!]
\includegraphics[width=1.0\textwidth,angle=0]{figures/plot3diagsData.png} 
\caption{An illustration of multiple stellar populations.
  The left panel is a color-magnitude diagram  for 841,000 stars from the SDSS Stripe 82 Standard Star Catalog
(variable sources are excluded, \citealt{2021MNRAS.505.5941T}) that have Gaia matches within 0.15 arcsec (after correcting for proper
motion using Gaia measurements). A subset of 415,000 stars with $r < 22$ and $u<22$ are overplotted as blue dots, and 409,000 of those that
also have $0.2 < g-i < 3.5$ (dominated by main-sequence stars and red giants) are overplotted as cyan dots. Finally, 63,000 stars that have
a signal-to-noise ratio for Gaia’s parallax measurements of at least
20 are shown as green dots (these stars can be used for the
calibration of luminosity-color relations). The same symbol color
scheme is used in the other two panels. The three yellow lines in the
middle panel show the stellar locus parametrization used by Green et al. (2014) for three values of metallicity (left to right): $[Fe/H] = -2, -1, 0$. In the right panel, the impact of metallicity on color-color tracks is negligible and all three are indistinguishable from each other. In the bottom of the middle panel, at $-0.5 < g-r < 0$, the three dark blue sequences correspond to (from left to right) He white dwarfs, H white dwarfs, and blue horizontal branch stars. The clouds of pale blue dots visible above the main stellar locus in the middle and right panels correspond to unresolved binary stars \citep{2004ApJ...615L.141S}.} \label{fig:3dataDiags}
\end{figure*}

Our approach adopted here is essentially the same as used in recent papers\footnote{This approach greatly simplifies when studying
faint and distant blue halo stars, as in e.g., \cite{2008ApJ...673..864J}. Such stars are beyond the dust layer which is confined close
to the disk and thus $A_r$ can be obtained from IR maps, they have halo metallicities ($[Fe/H] \sim -1.5$), and can be assumed dominated
by main-sequence stars. As a result, a simple functional relationship, $M_r = f(g-i)$, or its generalized version that
accounts for the shift of $M_r$ as a function of metallicity \citep{2008ApJ...684..287I}, can be used to estimate distance
in a straightforward manner.}
by, e.g., \cite{2011MNRAS.411..435B},  \cite{2014ApJ...783..114G}, \cite{lallement_3d_2014}, 
\cite{gordon_panchromatic_2016}, \cite{queiroz_starhorse_2018}, \cite{green_3d_2019} and \cite{bailer-jones_estimating_2021}.
The main differences compared to these works include: 
\begin{enumerate}
\item The use of multiple stellar populations, in addition to main-sequence stars and red giants (white dwarfs, unresolved binary stars,
                blue horizontal branch stars; potentially Miras, quasars and RR Lyrae stars too, which can also be recognized and rejected using variability).
\item Improved color tracks for main-sequence stars and (especially) red giants, including the use of very young ($<$1 Gyr) populations 
               and an extended $[Fe/H]$ range. 
\item Priors based on sophisticated TRILEGAL simulations by \cite{2022ApJS..262...22D} that include multiple stellar populations and
               also account for the Galaxy's bulge component (see
               \S\ref{sec:priors} below for more details). 
\end{enumerate}

@@

\begin{figure*}[b!]
\includegraphics[width=1.0\textwidth,angle=0]{figures/plot3HRdiags.png}  
\caption{The left panel shows SDSS-based empirical absolute magnitude vs. color parametrization for main-sequence stars and red giants.
The data are color-coded by metallicity, ranging from $[Fe/H] = -2.5$ to 0 (blue to red). The three lines correspond to three values of
metallicity: $[Fe/H] = -2, -1, 0$ (dot-dashed, solid and dashed, respectively). The middle panel shows a sample of 63,000 stars that have
signal-to-noise ratio for Gaia’s parallax measurements of at least 20 (white dwarfs can be seen in the lower left corner.
Their absolute magnitudes are derived from parallax measurements. The dot-dashed,
solid and dashed black lines are the same as in the left panel. For comparison, the essentially identical yellow dotted lines were computed
using eqs. A2 and A7 from Ivezi\'{c} et al. (2008).  Note the discrepancy between these parametrizations and data for sub-giant stars ($M_r \sim 3-4$
and $g-i \sim 0.8-1.1$). The right panel shows a sample of 415,000 stars with $r < 22$ and $u<22$ as red dots (shown by blue dots in
Figure~\ref{fig:3dataDiags}), and 409,000 of those that also have  $0.2 < g-i < 3.5$ as cyan dots. Their absolute magnitudes were computed
using the so-called “photogeometric” distances from Bailer-Jones et al. (2021). The dot-dashed, solid and dashed black lines are the
same as in the left and middle panels. About 10,000 stars below the main sequence (about 2.5\% of the full sample) seen at $g-i = 0.4$ and
$Mr > 7$ are predominantly found at the faint end ($r>20$) and may be outliers due to the low photometric signal-to-noise ratio.}
\label{fig:3HRdiags}
\end{figure*}

@@

We ``augment'' the empirical SDSS-based isochrones in two steps:
\begin{enumerate}
\item We extend its metallicity range from $[Fe/H]=0$ to $[Fe/H]=+0.5$ by linear extrapolation of the color vs. metallicity
  dependence around $[Fe/H]=0$. We adjust the gradient by a multiplicative factor (0.7) to ensure that the bright end of the
  stellar locus at $[Fe/H]=+0.5$ agrees with SDSS-Gaia data (shown in the middle panel in Figure~\ref{fig:3HRdiags}).
\item 
  For $M_r < M_{TO}$, where turn-off absolute magnitude $M_{TO}$ depends on age and ranges from $M_{TO}$=4 for age = 1 Gyr
  to $M_{TO}$ = 5 for age = 10 Gyr, we ``attach'' model-based
  isochrones (we use DSED isochrones hereafter) to the empirical
  SDSS-based isochrones.
\end{enumerate}

@@

We sample the M dwarf luminosity in the range $8.5 \le M_r \le 14.5$, with a step of 0.1 mag, and for each value
generate a track by adding the luminosity of a white dwarf, where the track is sampled on the same grid of $M_r$ but
using white dwarf models (separately for H and He models). There are four model families (for two families of white dwarf
models and two values of $[Fe/H]$ for M dwarfs), each with 3,721 $M_r$ vs. color entries. The color tracks for H white dwarfs
and $[Fe/H]=-1.5$ M dwarfs are shown in Figure~\ref{fig:locusBinaries}. Analogous tracks for binaries composed of an M dwarf with
$[Fe/H]=0$ and either an H or He white dwarf look very similar though not identical (differences are at most a few tenths
of a magnitude). In practice, it will be nearly impossible to constrain M dwarf metallicity, and very hard to distinguish
H and He white dwarfs. It may turn out that it will be sufficient to use a single model family to fit LSST data (e.g., a
combination of an H white dwarf and a $[Fe/H]=-1.5$ M dwarf).

@@

Studies, such as \cite{2012ApJ...757..166B}, based on SDSS data (including the Galactic plane), 
find that the scatter of $R_V$ around its mean value $R_V=3.1$ is very small. For this reason, we adopt
empirical results from their Table 1: $C_m$ = (1.810, 1.400, 0.759, 0.561) in $(u, g, i, z)$, respectively ($C_r=1$
by definition). This choice does not preclude the use of our framework to fit for $R_V$ as the fourth free
model parameter in Galactic plane regions with very large $A_r$ ($R_V$ is poorly constrained when $A_r$ is small);
we  can simply add multiple models with extincted colors generated using different values of $C_m$.

@@

In the next Section, we discuss a fast numerical pipeline implementation that can perform this computation for LSST-sized
catalogs and provide a more quantitative analysis of the method's performance.  We conclude this Section with a brief
discussion of the Bayesian model selection method for assigning posterior probabilities to each stellar population used to interpret
observations.
